---
title: "Example Data Files"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example Data Files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The `andorR` package includes several raw data files in different formats to demonstrate the data loading functions. These files can be found and loaded using the `system.file()` command.

### Included Files

```{r, echo=FALSE, results='asis'}
# This code chunk finds and lists the files for you
# NOTE: We use try() in case the package isn't fully loaded during a check
files <- try(list.files(system.file("extdata", package = "andorR")), silent = TRUE)

if (length(files) > 0 && !inherits(files, "try-error")) {
  cat("The following example data files are included in the `inst/extdata` directory:\n\n")
  cat(paste("*", files), sep = "\n")
} else {
  cat("No example data files were found in the package's `inst/extdata` folder.\n")
}
```

### File documentation

#### Ethical investiment decision tree (_ethical_)

#### Notifiable aquatic animal disease decision tree (_notifiable_)

#### Bovine Spongiform Encephalopathy surveillance decision tree (_bse_)

#### Multiple sclerosis diagnostic decision tree (_ms_)

#### Systemic lupus erythematous diagnostic decision tree (_lupus_)


### Example: Loading a File

To load the example YAML file from within the package, you would use the following code:
    
```{r, eval=FALSE}
path <- system.file("extdata", "ethical_investment.yml", package = "andorR")
my_tree <- load.tree.yaml(path)
```
